---
title: "PSTAT 120C HW 4"
author: "TJ Sipin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = T)

library(dplyr)
library(knitr)
library(kableExtra)
```

# Reading

### In your own words, compare and contrast parametric and nonparametric statistcs.

Parametric statistics are used when general assumptions hold. Nonparametric statistics are used in situations in which specific assumptions do not hold or are not particularly certain to be held. They can also be useful for when studying the effects on data that aren't intuitively mathematical, such as those that are subjective. 

Additionally, parametric statistics are used to apply to problems in which the the sample distributions are specified, except for the values of a finite number of parameters. Nonparametric statistics is used when they are not specified.

### What are two reasons why researchers might choose a nonparametric test over its parametric equivalent?

It may be useful if assumptions do not hold or may not hold, such as constant variance. They are likely to be more powerful in detecting population differences when the assumptions are not satisfied.

Additionally, nonparametric statistics can be used for observations that are difficult to quantify.

### List the nonparametric tests we’ve discussed that are appropriate for paired samples.

+ Sign test
+ Wilcoxon signed-rank test

### List the nonparametric tests we’ve discussed that are appropriate for independent samples.
+ Sign test
+ Wilcoxon signed-rank test
+ Mann-Whitney U
+ Kruskal-Wallis test
+ Friedman test

### What assumptions do most nonparametric tests appear to make?

+ Independence
+ Number of measurements in the sample from population $i$ must be greater than or equal to 5.

# Practice

1. The following data record the number of ear infections each of 7 swimmers had before ($X_i$) taking a medication designed to prevent infections and after ($Y_i$) taking the medication. Using the data provided, answer the following questions:

  | Swimmer | $X_i$ | $Y_i$ | $D_i$ | Sign |
  |:---:|:---:|:---:|:---:|:---:|
  | A | 3 | 2 |  |  |
  | B | 0 | 1 |  |  |
  | C | 5 | 4 |  |  |
  | D | 4 | 0 |  |  |
  | E | 2 | 1 |  |  |
  | F | 4 | 3 |  |  |
  | G | 3 | 1 |  |  |
  
(a) Consider $D_i = X_i - Y_i$. Fill in the $D_i$ column. Then fill in the "Sign" column with the sign of the differences.
  
  | Swimmer | $X_i$ | $Y_i$ | $D_i$ | Sign |
  |:---:|:---:|:---:|:---:|:---:|
  | A | 3 | 2 | 1 | + |
  | B | 0 | 1 | -1 | - |
  | C | 5 | 4 | 1 | + |
  | D | 4 | 0 | 4 | + |
  | E | 2 | 1 | 1 | + |
  | F | 4 | 3 | 1 | + |
  | G | 3 | 1 | 2 | + |
  
(b) Use the *sign test* to conduct a hypothesis test assessing whether there is a difference between the population distributions of $X$ and $Y$.
  
  - Write the null and alternative hypotheses in words and in symbols.;
  
    $H_0: p = 0.50,$ where $p = P(X > Y)$, or it is equally likely for $X$ to be greater than $Y$ as it is for $Y$ to be greater than $X$.
  
  - Calculate a test statistic $M$;
  
    $M = $ # of positive differences where $D_i = X_i - Y_i$.
    
    $$
    M = 6
    $$
    
  - Calculate the relevant binomial probability;
    $$
    \begin{align}
    P(M \ge 6) &= {7 \choose 6} 0.5 ^6 0.5 ^1 + {7 \choose 7} 0.5^7 0.5 ^0 \\
    &= 7(0.5^7) + 1(0.5^7) \\
    &= 0.0625 \\
    &= \frac{1}{16} \\
    P(M \le 1) &= {7 \choose 1}0.5^1 0.5^6 + {7 \choose 0} 0.5^7 \\
    &= 7(0.5^7) + 0.5^7 \\
    &= 0.0625 \\
    &= \frac{1}{16}
    \end{align}
    $$
    
    So the two-tailed p-value for $M = 6, n = 7$ is $\frac{1}{8} = 0.125.$
    
  - Can you reject the null hypothesis, assuming $\alpha = 0.05$?
  
    Since $p = 0.125$, we would fail to reject since $p > 0.05$.
    
(c) Use the Wilcoxon signed-rank test to conduct a hypothesis test assessing whether there is a difference between the population distribution $X$ and $Y$.
  
  - Calculate a test statistic $T$;
  
    $$
    \begin{align}
    T^+ &= 1.2 + 1.2 + 7 + 1.2 + 1.2 + 6 = 17.8 \\
    T^- &= 1.2 \\
    T &= \min(17.8, 1.2) = 1.2
    \end{align}
    $$
    
  - Determine the appropriate critical value for $\alpha = 0.05$ (or as close as possible to 0.05);
  
    Since $n = 7$, our critical value is 2.
    
  - Can you reject the null hypothesis?
  
    We reject if $T \leq T_0 = 2$. Since $T = 1.2$, then we reject the null hypothesis and conclude that there is a difference in ear infections among swimmers before taking medication and after.

(d) Were you able to reject the null with one test, both tests, or neither? If you rejected with one test and not the other, what do you think contributed to this difference?

  We rejected the null hypothesis for the Wilcoxon signed-rank test. For the sign test, we are looking purely at the sign of the difference, rather than the ranks, so there is a sense of generality and less information for the former. This may have led to less inferential power, leading us to fail to reject since we lacked evidence to reject, as opposed to the Wilcoxon signed-rank test.
  
2. The coded values for a measure of brightness in paper (light reflectivity), prepared by two different processes, are as shown in the accompanying table for samples of size **9** drawn randomly from each of the two processes. Do the data present sufficient evidence to indicate a difference in locations of brightness measurements for the two processes? Give the attained significance level.

| **Process A** | **Process B** |
|---------------|---------------|
| 6.1           | 9.1           |
| 9.2           | 8.2           |
| 8.7           | 8.6           |
| 8.9           | 6.9           |
| 7.6           | 7.5           |
| 7.1           | 7.9           |
| 9.5           | 8.3           |
| 8.3           | 7.8           |
| 9.0           | 8.9           |

  a. Answer the question by using the **Mann-Whitney U test**.
  
  



```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
